\section{A Crash Course on First-Order Logic}

We begin by recalling the notion of a first-order language.

\begin{boxdefinition}[Language]
    A \textbf{language} is a disjoint union
    \begin{align*}
        \Lang = \Funcs \cup \Rels \cup \Consts
    \end{align*}
    of countable sets of symbols, where
    \begin{itemize}
        \item $\Funcs$ is a set of function symbols
        \item $\Rels$ is a set of relation symbols
        \item $\Consts$ is a set of constant symbols
    \end{itemize}
\end{boxdefinition}

Next, we recall the notion of a structure in a language.

\begin{boxdefinition}[Structure]
    Let $\Lang$ be a structure. An \textbf{$\Lang$-structure} is a tuple
    \begin{align*}
        M = \cycl{U; F, R, C}
    \end{align*}
    consisting of a non-empty set $U$ and functions, relations, and constants that live in $\Funcs$, $\Rels$ and $\Consts$ respectively. $U$ known as the \textbf{universe} of $M$, and is denoted by $\abs{M}$. The function, relation and constant symbols of $M$ are denoted $F^M$, $R^M$ and $C^M$ respectively.
\end{boxdefinition}

Any function or relation in a structure has an \textbf{arity}, which is informally the number of arguments it takes. An important fact to note is that arities are not a feature of functions and relations themselves, but of their corresponding \textit{symbols}. In other words, \textbf{arity is a syntactic notion}. Semantically speaking, when we seek an interpretation of a function symbol of some arity $n$, we are forced to limit our search to the set of functions from $U^n$ to $U$.

We now describe the notion of structure-preserving bijections, known as isomorphisms.

\begin{boxdefinition}[Isomorphism]
    Let $\Lang$ be a language and let $M, N$ be $\Lang$-structures. We say that a function $g : \abs{M} \to \abs{N}$ is an \textbf{isomorphism} if
    \begin{enumerate}
        \item $g$ is a bijection
        \item ``$g$ commutes with functions''
        \item ``$g$ commutes with relations''
    \end{enumerate}
    where the double-quotes for the second and third point above refer to the fact that we implicitly require an equality of arities condition before we can talk about composing isomorphisms with multi-ary functions.
\end{boxdefinition}

% Lecture 2

We are now ready to define submodels.

\begin{boxdefinition}[Submodel]
    Let $M, N$ be $\Lang$-structures. We say that $M$ is a submodel of $N$, denoted $M \susbeteq N$, if
    \begin{enumerate}
        \item $\abs{M} \subseteq \abs{N}$.
        \item For all function symbols $F\of{x_1, \ldots, x_n}$, the interpretation in $M$ agrees with the interpretation in $N$.
        \item For all relation symbols $R\of{a_1, \ldots, a_n}$, the interpretation in $M$ agrees with the interpretation in $N$.
        \item For all constant symbols $C$, the interpretation in $M$ agrees with the interpretation in $N$.
    \end{enumerate}
\end{boxdefinition}

In particular, a submodel of a model is also a model. For instance, if $G$ is a group, ie, a model of the group axioms, then any submodel of $G$ is, in fact, a subgroup of $G$, and a group in its own right (in that it again models the group axioms).

We now talk about more syntactic elements of a language.

\begin{boxdefinition}[Terms]
    Let $\Lang$ be a language. $\Term(\Lang)$ is defined to be the minimal set of finite sequences of symbols\footnote{Here, the set of variable symbols is countable, but we might, on occasion, need uncountably many variable symbols} from
    \begin{align*}
        \set{(, ), [, ]} \cup \Consts \cup \Funcs \cup \set{x_1, x_2, x_3, \ldots}
    \end{align*}
    satisfying the following rules:
    \begin{enumerate}
        \item Every constant symbol is a term.
        \item Every variable is a term.
        \item For all $n$-ary functions $f$ and $n$ terms $t_1, \ldots, t_n$, $f\of{t_1, \ldots, t_n}$ is a term.
        \item Every term arises in this way.
    \end{enumerate}
\end{boxdefinition}

% For an ULTRA-FORMAL presentation of the material, look at D. Monk Mathematical Logic, Springer GTM. But Rami's feedback is: an undergraduate cannot follow it because it is too technical, and a graduate student will find that it has too much detail but not much actual content.

In similar fashion, we can define the formulae in a language.

\begin{boxdefinition}[Formulae]
    Let $\Lang$ be a language. $\Fml(\Lang)$ is defined to be the minimal set of finite sequences of symbols from
    \begin{align*}
        \Term\of{L} \cup \set{\land, \lor, \to, \neg, \ldots} \cup \set{=} \cup \set{\forall, \exists}
    \end{align*}
    satisfying the following rules:
    \begin{enumerate}
        \item For all $\tau_1, \tau_2 \in \Term\of{\Lang}$, $\tau_1 = \tau_2$ is a formula.
        \item For all $n$-ary relations $R$ and $n$ terms $t_1, \ldots, t_n$, $R\of{t_1, \ldots, t_n}$ is a formula.
        \item For all connectives $\star$ and formulae $\Phi$ and $\Psi$, $\Phi \star \Psi$ is a formula.
        \item For a quantifier $Q$, variable $x$ and formula $\varphi(x)$, $Qx \parenth{\varphi\of{x}}$ is a formula.
        \item Every formula arises in this way.
    \end{enumerate}
    Formulae consisting only of a single relation symbol (including formulae that only consist of an equality) are called \textbf{atomic formulae}. The atomic formulae of $\Lang$ are denoted $\AFml(\Lang)$.
\end{boxdefinition}

For a formula $\varphi$, denote by $\FV{\varphi}$ the set of free variables of $\varphi$. It is sometimes useful to distinguish those formulae in a language that contain no free variables.

\begin{boxdefinition}
    Define the set of \textbf{satisfiable} formulae of a language $\Lang$ to be
    \begin{align*}
        \Sat\of{\Lang} := \setst{\varphi \in \Fml\of{\Lang}}{\FV{\varphi} = \emptyset}
    \end{align*}
\end{boxdefinition}

Finally, we make the most important definition of this section.

\begin{boxdefinition}[Satisfaction]
    Let $\Lang$ be a language and $M$ an $\Lang$-structure. For all $\varphi \in \Fml\of{\Lang}$, we say that \textbf{$M$ models $\varphi$}, denoted $M \models \varphi$, if \sorry % This is soooo tedious fr
\end{boxdefinition}

We are now ready to state a simple-sounding but rather non-trivial result.

\begin{boxlemma}
    Suppose $M$ and $N$ are both $\Lang$-structures. If $M \subseteq N$, then for all $\tau \in \Term\of{\Lang}$, $\tau^{M}\!\brac{a} = \tau^N\!\brac{a}$, where $a \in \abs{M} \times \cdots \times \abs{M}$ and $\tau^M\!\brac{a}$ and $\tau^N\!\brac{a}$ denote interpretations of $\tau$ in $M$ and $N$ with the variables all being interpreted as the components of $a$.
\end{boxlemma}

We do not prove this result. It is not difficult.

Next is a less trivial result.

\begin{boxlemma}
    If $M \subseteq N$, then $M \models \varphi$ if and only if $N \models \varphi$ for all quantifier-free formulae $\varphi$.
\end{boxlemma}