\section{Models of Peano Arithmetic}

In this section, we build many models of Peano arithmetic, the most standard and universal of which is $\omega$, the natural numbers.

\subsection{Restrictions and Expansions}

Before going any further, we will need to define a central tool: restrictions and expansions. Throughout this subsection, fix a language $L$ and an $L$-structure $M$.

\begin{boxdefinition}[Restriction/Expansion]
    Let $L_1 \subseteq L$, so that $L$ contains relations, functions, and constants $\Rels\of{L_1}$, $\Funcs\of{L_1}$, and $\Consts\of{L_1}$. The \textbf{restriction of $M$ to $L$} is the $L_1$-structure
    \begin{align*}
        M \vert_{L_1} := \cycl{\abs{M}, \Rels^{M}\of{L_1}, \Funcs^M\of{L_1}, \Consts^M\of{L_1}}
    \end{align*}
    Dually, we say that \textbf{$M$ is the expansion of $M\vert_{L_1}$ to $L$}.
\end{boxdefinition}

A good example of this is to model-theoretically encode the fact that every field is also an abelian group (additively).

\begin{boxexample}[Restriction: Fields to Abelian Groups]
    Let $L$ be the language of fields and let $L_1$ be the language of (additively expressed) abelian groups. Then, if
    \begin{align*}
        M = \cycl{\Q, +, \cdot, 0, 1}
    \end{align*}
    then its restriction to $L_1$ is
    \begin{align*}
        M\vert_{L_1} = \cycl{\Q, +, 0}
    \end{align*}
\end{boxexample}

\subsection{The Language and Theory of Peano Arithmetic}

In this subsection, we set up our study of Peano arithmetic by defining the language in which we will work and the theory we will seek to model. We begin with some notation.

\begin{boxnotation}
    Let 
\end{boxnotation}

\subsection{Existence of Non-Standard Models of Peano Arithmetic}

We know that there is a standard model of Peano arithmetic, denoted $N$. This consists intuitively of the natural numbers---that is, ordinals less than $\omega$---with the usual addition, multiplication, additive and multiplicative identities. It turns out there are also other models, whose existence we will see in this subsection.

Before that, we mention that any model of Peano arithmetic admits a linear order.

\begin{boxtheorem}
    The following is a valid deduction.
    \begin{align*}
        \PA \vdash \forall x \forall y \exists z \parenth{y = x + z}
    \end{align*}
\end{boxtheorem}

\begin{boxtheorem}
    There exists some $M\models \PA$ such that $M$ is not isomorphic to the standard model $N$.
\end{boxtheorem}
\begin{proof}
    Augment the language $L$ to the $L_1 = L_{\PA} \cap \set{c}$ by adding a new constant symbol $c$. Let
    \begin{align*}
        T_1 := \PA \cup \set{c \neq 0} \cup \setst{c \neq \underbrace{1 + \cdots + 1}_{n \text{ times}}}{n < \omega}
    \end{align*}
    We show, using the Compactness Theorem (\sorry), that there is a model $M$ of $T_1$. Given $T_0 \subseteq T_1$ finite, let $n_0 < \omega$ be the largest natural number such that
    \begin{align*}
        c \neq \underbrace{1 + \cdots + 1}_{n \text{ times}}
    \end{align*}
    lies in $T_0$. Define the $L_1$-structure
    \begin{align*}
        M_0 := \cycl{\omega, +, \cdot, 0, 1, a}
    \end{align*}
    be the expansion of $N$ to $L_1$., with $c^{M_0} = a$. Then, $M_0 \models T_0$. Hence, since every finite subset of $T_1$ has a model, so does $T_1$. Call this model $M_1$. Denote by $M$ the restriction of $M_1$ to $L_{\PA}$.

    Now that we have established the existence of $M$, all that remains is to show that $M$ is not isomorphic to $N$. Let $b = c^M$. Suppose $M \cong N$ via an isomorphism $f$. Then, $f(b) \in \omega$, so there is some $n < \omega$ such that
    \begin{align*}
        f(b) = \underbrace{1 + \cdots + 1}_{n \text{ times}}
    \end{align*}
   
\end{proof} % I'm slightly unsure about interpreting c in M as b, because c is a symbol in the larger language L_1, not L... I get what you're going for (makes a lot of sense) but I'm stuck on the previous step - defining b. Will maybe ask Rami after class

%asking would be good - as I understand, all we need for the prooof is that there is an interpretation of c in M (c arises as a constant in the larger language, but nonetheless the existence of an interpretation implies that such an element exists in M) - I think now he's working on something else

% Makes sense. Yeah this might be something else

%we can check after

% (Unless you know?)
%(lmao)
% Oh wait. His proof isn't over

\subsection{Countable Models of Peano Arithmetic}

Recall the definition of the \textit{spectrum of a theory} (\Cref{Ch1:Def:Spectrum}). It turns out, we can use the idea of a spectrum to say something rather sophisticated about countable models of Peano Arithmetic.

\begin{boxtheorem}
    $I\of{\aleph_0, \PA} = 2^{\aleph_0}$.
\end{boxtheorem}

\subsection{Impossible Deductions}

We know that $\PA$ admits a model (indeed, a standard model). Therefore, we know that $\PA$ is a consistent theory. However, Gödel famously showed that it is not possible to prove consistency of $\PA$ using $\PA$ alone. That is, he proved the following.

\begin{boxtheorem}[Gödel's Second Incompleteness Theorem]
    $\PA \not\vdash \PA \text{ is consistent}$
\end{boxtheorem}
\begin{remark}
    Note that no number theorist would accept this as an example of a statement of interest in number theory which is not provable in $\PA$ (so says the model theorist). 
\end{remark}

Another interesting impossibility result involves Ramsey theorem.

\begin{boxtheorem}[Paris-Harrington 1976]
    $\PA \not\vdash \text{A special case of Ramsey's Theorem}$.
\end{boxtheorem}

% I'm doing this elsewhere

% Follow me